---
title: "BMR staging with ALSFRS revised scores"
output: html_notebook
---

```{r message=FALSE}
library(tidyverse)
library(purrr)
```

```{r}
source("helper.R")
```

### Grouping ALSFRS 12 items into 3 domains  
Q1-3 -> Bulbar 
Q4-10 -> Motor 
Q11-12 -> Respiratory   

### Staging for each domain   
**general definition**  
stage 0; not affected, normal 
stage 1; affected 
stage 2; loss of autonomy 

**staging for bulbar domain**   
if both Q1 and Q3 item scores are 4, then assign stage 0. (note that we exclude Q2 item)     
if either Q1 or Q3 item score is 1 or less, then assign stage 2.   
otherwise assign stage 1.  

**staging for motor domain**    
if all item scores of motor domain (Q4-8) are 4, then assign stage 0. (note that we exclude Q9)   
if two or more item scores are 1 or less, then assign stage 2.
otherwise assign stage 1.  

**staging for respiratory domain**    
if all respiratory items (Q10-12) scores are 4, then assign stage 0.   
if any item score is 1 or less, then assign stage 2.   
otherwise assign stage 1.  


```{r}
# read in alsfrs revised data
alsfrs = read.csv("~/Dropbox/ALSmaster/PROACT/ALSFRS_rev.csv")
names(alsfrs) = tolower(names(alsfrs))
```


it may take long to execute the following code bloc 
```{r}
# calculate b_m_r_stage from alsfrs item scores 
alsfrs_domain = alsfrs %>% 
  group_by(subjectid, feature_delta) %>%
  nest() %>%
  mutate(b_stage = map_dbl(data, ~convert_bulbar(.x[c("q1_speech","q3_swallowing")]))) %>%
  mutate(m_stage = map_dbl(data, ~convert_motor(.x[c("q4_handwriting","q5_cutting","q6_dressing_and_hygiene","q7_turning_in_bed","q8_walking")]))) %>%
  mutate(r_stage = map_dbl(data, ~convert_resp(.x[c("r1_dyspnea","r2_orthopnea","r3_respiratory_insufficiency")])))
```

```{r}
head(alsfrs_domain)
```


```{r}
bmr = alsfrs_domain %>%
  unnest(data) %>%
  select(subjectid, feature_delta, 
         b_stage, m_stage, r_stage)
head(bmr)
```


```{r}
stage = bmr %>%
  mutate(bmr_stage = b_stage + m_stage + r_stage)
```


```{r}
surv = read.csv("data/survival.csv")
names(surv) = tolower(names(surv))
deceased = surv %>%
  filter(status == 1) %>%
  filter(subjectid %in% stage$subjectid) %>%
  mutate(bmr_stage = 7) %>%
  rename(feature_delta = time_event) %>%
  select(-status)
```


```{r}
stage_final = bind_rows(stage, deceased)
```


In some patients, the time of death coincided with the time of alsfrs recording.  
```{r}
stage_final %>%
  group_by(subjectid, feature_delta) %>%
  count() %>%
  filter(n > 1) -> temp
temp$subjectid -> ids 
temp$feature_delta -> deltas
```


```{r}
stage_final %>%
  filter(subjectid %in% ids) %>%
  filter(feature_delta %in% deltas)
```


```{r}
temp = stage_final %>%
  filter(subjectid %in% ids) %>%
  filter(feature_delta %in% deltas) %>%
  filter(!(bmr_stage == 7))
```


```{r}
stage = stage_final %>%
  anti_join(temp, by=c("subjectid","feature_delta","bmr_stage"))
```


```{r}
stage = within(stage, {
  b_stage = factor(b_stage)
  m_stage = factor(m_stage)
  r_stage = factor(r_stage)
  bmr_stage = factor(bmr_stage)
})
summary(stage)
```

```{r}
write_csv(stage, "~/Dropbox/ALSmaster/PROACT/stage_bmr.csv")
```


