---
title: "ALS staging; EDA"
output: html_notebook
---

```{r}
library(tidyverse)
library(broom)
library(gridExtra)
```


```{r}
# Read stage dataset 
stage1 <- read_csv("data/stage_king_mitos.csv")
stage2 <- read_csv("data/stage_bmr.csv")
```

```{r}
stage = stage1 %>%
  inner_join(stage2, by = c("subjectid", "feature_delta"))
```

```{r}
stage = within(stage, {
  king = factor(king)
  mitor = factor(mitos)
  b_stage = factor(b_stage)
  m_stage = factor(m_stage)
  r_stage = factor(r_stage)
  bmr_stage = factor(bmr_stage)
})
```



```{r}
# read in alsfrs_rev data 
```


```{r}
# merge alsfrs_rev and stage dataframes 
```

time taken from symptom onset to reach each stage (last recorded for corresponding stage, every records for corresponding stage)  

standardized time taken to reach each stage (time taken to reach each stage, divided by entire disease duration, ie., from symptom onset to death; 0, symptom onset; 1, death)  

some patients may have stage reversal. 

comparisons btw bulbar onset vs. limb onset.   

effect of riluzole use on time taken to reach each stage  


KM curve analysis; from last recorded stage to death or censor date 













