---
 titie: "Changes of stages with disease progression"
 author: "Ilhan Yoo"
 date: "r format(Sys.Date())"
 output: word_document: 
 fig_height: 4
 fig_width: 4
 toc: no
 ---
 
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE,fig.align="center",message=F,warning=F,dpi=300,dev="tiff")
```



# library loading
```{r}
library(tidyverse)
library(gridExtra)
```



# BMR stage dataset
# King and MiToS stage dataset, subjectid의 이름은 bmr stage dataset과 같은 SubjectID로 변경 
```{r}
bmr <- read.csv("stage_bmr.csv")
king_mitos <- read.csv("stage_king_mitos.csv") %>% rename(SubjectID=subjectid)
```



# distribution of ALS stage last recorded in 12 month period by baseline stage, exclude patients with fu period of less than 9 months 
# bmr_feature_delta_12mo, king_mitos_feature_delta_12mo: f/u data가 9~15개월의 데이터가 있는 대상자만 filtering
```{r}
bmr_feature_delta_12mo <- bmr %>% group_by(SubjectID) %>% 
  mutate(first_bmr=first(bmr_stage)) %>% 
  ungroup() %>% 
  filter(between(feature_delta,9,15))
king_mitos_feature_delta_12mo <- king_mitos %>% group_by(SubjectID) %>% 
  mutate(first_king=first(king),first_mitos=first(mitos)) %>%
  ungroup() %>% 
  filter(between(feature_delta,9,15)
```



# bmr_feature_delta_12mo, king_mitos_feature_delta_12mo: f/u data 9~15개월의 데이터 중 12개월에 가장 근접한 데이터만 추출 
```{r}
bmr_feature_delta_12mo <- bmr_feature_delta_12mo %>% group_by(SubjectID) %>% 
  mutate(temp_feat=abs(feature_delta-12)) %>% 
  filter(temp_feat==min(temp_feat)) %>% 
  dplyr::select(-temp_feat)
king_mitos_feature_delta_12mo <- king_mitos_feature_delta_12mo %>% group_by(SubjectID) %>% 
  mutate(temp_feat=abs(feature_delta-12)) %>% 
  filter(temp_feat==min(temp_feat)) %>% 
  dplyr::select(-temp_feat)
```



# baseline stage에 따라서 12개월쯤(9~15개월 사이 f/u데이터 중 12개월에 가장 근접한 f/u 시점)의 stage변화
```{r}
p_bmr <- bmr_feature_delta_12mo %>% ggplot(aes(x=factor(first_bmr),fill=factor(bmr_stage)))+
  geom_bar(position="dodge")+
  theme_bw()+
  labs(title="distribution of BMR stage after 12 months",x="Baseline BMR stage",y="Number of subject",fill="BMR stage after \n about 12 months")+
  theme(panel.grid.major = element_blank(),panel.grid.minor = element_blank(),plot.title = element_text(hjust = 0.5))
p_king <-king_mitos_feature_delta_12mo %>% ggplot(aes(x=factor(first_king),fill=factor(king)))+
  geom_bar(position="dodge")+
  theme_bw()+
  labs(title="distribution of King's stage after 12 months",x="Baseline King's stage",y="Number of subject",fill="King's stage after \n about 12 months")+
  theme(panel.grid.major = element_blank(),panel.grid.minor = element_blank(),plot.title = element_text(hjust = 0.5))
p_mitos <-king_mitos_feature_delta_12mo %>% ggplot(aes(x=factor(first_mitos),fill=factor(mitos)))+
  geom_bar(position="dodge")+
  theme_bw()+
  labs(title="distribution of MiToS stage after 12 months",x="Baseline MiToS stage",y="Number of subject",fill="MiToS stage after \n about 12 months")+
  theme(panel.grid.major = element_blank(),panel.grid.minor = element_blank(),plot.title = element_text(hjust = 0.5))
```



# merge 3 plots of each stage system
```{r}
grid.arrange(p_bmr,p_king,p_mitos,ncol=1)
```

