---
 titie: "Changes of stages with disease progression"
 author: "Ilhan Yoo"
 date: "r format(Sys.Date())"
 output: word_document: 
 fig_height: 4
 fig_width: 4
 toc: no
 ---
 
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE,fig.align="center",message=F,warning=F,dpi=300,dev="tiff")
```



# library loading
```{r}
library(tidyverse)
library(gridExtra)
```



# 이전시점과 다음 방문 시점의 BMR stage를 비교하여 transition probability table 만듦 
```{r}
bmr_tran <- bmr %>% group_by(SubjectID) %>% 
  mutate(bmr_stage_next=lead(bmr_stage)) %>% 
  filter(!is.na(bmr_stage_next))
bmr_tran_mtrx_prop <- prop.table(table(bmr_tran$bmr_stage,bmr_tran$bmr_stage_next),2)

king_tran <- king_mitos %>% group_by(SubjectID) %>% 
  mutate(king_stage_next=lead(king)) %>% 
  filter(!is.na(king_stage_next))
king_tran_mtrx_prop <- prop.table(table(king_tran$king,king_tran$king_stage_next),2)

mitos_tran <- king_mitos %>% group_by(SubjectID) %>% 
  mutate(mitos_stage_next=lead(mitos)) %>% 
  filter(!is.na(mitos_stage_next))
mitos_tran_mtrx_prop <- prop.table(table(mitos_tran$mitos,mitos_tran$mitos_stage_next),2)
```



# transition probability transformation: matrix to dataframe form 
```{r}
bmr_tran_mtrx_prop1 <- melt(bmr_tran_mtrx_prop)
king_tran_mtrx_prop1 <- melt(king_tran_mtrx_prop)
mitos_tran_mtrx_prop1 <- melt(mitos_tran_mtrx_prop)
```



# transition probability plot with confusion matrix form
```{r}
p_bmr <- bmr_tran_mtrx_prop1 %>% ggplot(aes(x=Var1,y=Var2,fill=value))+
  geom_raster()+
  geom_text(aes(label=round(value,2)))+ 
  labs(x="BMR stage",y="BMR stage of next visit",fill="Transition probability")+
  scale_fill_viridis_c()+
  scale_x_continuous(breaks=seq(0,7,by=1))+
  scale_y_continuous(breaks=seq(0,7,by=1))+
  theme(panel.grid.major = element_blank(),panel.grid.minor = element_blank())
p_king <- king_tran_mtrx_prop1 %>% ggplot(aes(x=Var1,y=Var2,fill=value))+
  geom_raster()+
  geom_text(aes(label=round(value,2)))+ 
  labs(x="King's stage",y="King's stage of next visit",fill="Transition probability")+
  scale_fill_viridis_c()+
  scale_x_continuous(breaks=seq(0,5,by=1))+
  scale_y_continuous(breaks=seq(0,5,by=1))+
  theme(panel.grid.major = element_blank(),panel.grid.minor = element_blank())
p_mitos <- mitos_tran_mtrx_prop1 %>% ggplot(aes(x=Var1,y=Var2,fill=value))+
  geom_raster()+
  geom_text(aes(label=round(value,2)))+ 
  labs(x="MiToS stage",y="MiToS stage of next visit",fill="Transition probability")+
  scale_fill_viridis_c()+
  scale_x_continuous(breaks=seq(0,5,by=1))+
  scale_y_continuous(breaks=seq(0,5,by=1))+
  theme(panel.grid.major = element_blank(),panel.grid.minor = element_blank())
```



# merge 3 plots of each stage system
```{r}
grid.arrange(p_bmr,p_king,p_mitos,ncol=1)
```

