---
 titie: "Changes of stages with disease progression"
 author: "Ilhan Yoo"
 date: "r format(Sys.Date())"
 output: word_document: 
 fig_height: 4
 fig_width: 4
 toc: no
 ---
 
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE,fig.align="center",message=F,warning=F,dpi=300,dev="tiff")
```



# library loading
```{r}
library(tidyverse)
library(gridExtra)
```



# BMR stage dataset
# King and MiToS stage dataset, subjectid의 이름은 bmr stage dataset과 같은 SubjectID로 변경 
```{r}
bmr <- read.csv("stage_bmr.csv")
king_mitos <- read.csv("stage_king_mitos.csv") %>% rename(SubjectID=subjectid)
```



# standardised time from symptom onset to each stage for only deceased patients 
# alsfrs_death: deceased patient dataset
```{r}
surv <- read.csv("survival.csv")
clinical_info <- read.csv("PROACT_preprocessed.csv")
death <- surv %>% filter(status==1)
```



# clinical info정보 import, onset으로부터 경과시간 계산 
```{r}
bmr_death <- alsfrs_bmr %>% filter(SubjectID %in% death$SubjectID) %>% merge(clinical_info, all.x = T, by="SubjectID")
bmr_death <- bmr_death %>% mutate(time_from_onset=round(abs(onset_delta)/30,2)+feature_delta,
                                        time_from_diag=round(abs(diag_delta)/30,2)+feature_delta)
king_mitos_death <- king_mitos %>% filter(SubjectID %in% death$SubjectID) %>% merge(clinical_info, all.x = T, by="SubjectID")
king_mitos_death <- king_mitos_death %>% mutate(time_from_onset=round(abs(onset_delta)/30,2)+feature_delta,
                                        time_from_diag=round(abs(diag_delta)/30,2)+feature_delta)
```



# standardised time 계산 & stage가 death인 경우 standardized time이 1이 아닌 경우는 제외 
```{r}
bmr_death <- bmr_death %>% group_by(SubjectID) %>% 
  mutate(stan_time=time_from_onset/max(time_from_onset)) %>% 
  arrange(SubjectID,feature_delta)
temp <- bmr_death %>% filter(bmr_stage==7&stan_time<1)
bmr_death <- setdiff(bmr_death,temp)

king_mitos_death <- king_mitos_death %>% group_by(SubjectID) %>% 
  mutate(stan_time=time_from_onset/max(time_from_onset)) %>% 
  arrange(SubjectID,feature_delta)
temp <- king_mitos_death %>% filter(king==5&stan_time<1)
king_mitos_death <- setdiff(king_mitos_death,temp)
```



# standardised median time plot, subject별로 BMR stage내에서 stage변하기 직전의 time from onset으로 standardized time계산
# BMR stage별 standardized median time plot 
```{r}
p_bmr <- bmr_death %>% group_by(SubjectID,bmr_stage) %>% 
  filter(feature_delta==max(feature_delta)) %>% 
  ungroup() %>% 
  ggplot(aes(factor(bmr_stage),stan_time,fill=factor(bmr_stage)))+
  geom_boxplot()+
  theme(panel.grid = element_blank())+
  coord_flip()+
  labs(x="BMR stage",y="Standardized median time",fill="BMR stage")
p_king <- king_mitos_death %>% group_by(SubjectID,king) %>% 
  filter(feature_delta==max(feature_delta)) %>% 
  ungroup() %>% 
  ggplot(aes(factor(king),stan_time,fill=factor(king)))+
  geom_boxplot()+
  theme(panel.grid = element_blank())+
  coord_flip()+
  labs(x="King's stage",y="Standardized median time",fill="King's stage")
p_mitos <- king_mitos_death %>% group_by(SubjectID,mitos) %>% 
  filter(feature_delta==max(feature_delta)) %>% 
  ungroup() %>% 
  ggplot(aes(factor(mitos),stan_time,fill=factor(mitos)))+
  geom_boxplot()+
  theme(panel.grid = element_blank())+
  coord_flip()+
  labs(x="MiToS stage",y="Standardized median time",fill="MiToS stage")
```



# merge 3 plots of each stage system
```{r}
grid.arrange(p_bmr,p_king,p_mitos,ncol=1)
```

